---
title: "Party Ethnicization, Partisan Affect, and Politial Trust"
author: "Ahmad S. Qabazard"
date: "4/8/2022"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Research Questions: 

Does ethnic voting influence the winner-loser gap in political trust?

## ELF and Democracy

- Previous work tends to use Ethno-Linguistic Fractionalization (ELF) to understand the effects of politicized ethnicity on democracy and development (Weber et al. 2016).  

- Findings on the relationship between ELF and democracy are mixed (See Gerring et al. 2016).  

- Ethnicization of party support as a better measure of politicized ethnicity (Huber 2012).  

## Ethnic Voting and Democracy:

- Reduces the quality of democracy by increasing the winner-take-all character of elections (Houle 2018).  
-- Argument focuses on elite behavior.  

- Dowd and Michael (2007) find consistent effects on alternative quality of democracy measure.  
-- Findings also show that ethnic voting lowers trust in government. 

## Political Trust

- Trust is the likelihood that institutions or their agents will produce preferred outcomes (Gamson 1958).
- Considered a diffuse form of support (Easton 1975; Van der Meer and Zmerli 2017). 
- Winner-loser status determines levels of support and this shows both cognitive and affective dimensions in trust judgments (Anderson and Guillory 1997).  
--Undermines democratic accountability.   
--Threatens democratic stability. 

## Ethnic Voting Increases Winner-Loser Gap in Political Trust

- Undermines the importance of cross-cutting cleavages (Lipset and Rokkan 1967).  
- Similar to winner-loser gap in trust, ethnic voting undermines democratic accountability (Carlson 2015). 
- Preferred outcomes (crucial for trust judgements) can be influenced by patronage distribution. Extending the winner-take-all character to individuals. 

## Hypothesis 1

- As party ethnicization increases, losers are less likely to trust political institutions compared to winners.  

## Congnitive Appraisal and Emotions 

- Lazarus and Folkman (1984).  
- Adopted to explain the role of emotions in literature on American partisanship (Mason 2016).  
- Allows for understanding the affective dimension of trust judgments.  
- Losing increases stress appraisal and, if stress causing object is identified, theory expects the production of anger as an emotional response.  
    -- Emphasis on losing (Anderson et al. 2005).  
    -- What determines the direction of emotional responses?  

## Hypothesis 2

- As party ethnicization increases, individuals are expected to exhibit higher levels of affective polarization.  

## Data and Methods: 

- Individual level:  
-- H1: Afrobarometer (Round 2 & 3).  
-- H2: CSES (Two Waves)

- Ordinary Least Squares (OLS) model with random effects.  Individuals nested in country-waves. 

## Dependent Variables:

- H1 Political trust: Index of trust in particular institutions.  
-- President, Parliament, Courts, Electoral Institutions, and Police.  
- H2 Affective polarization: Wagner (2020)'s Weighted Affective Polarization (WAP).  
-- Spread of party like-dislike scores using CSES items.  

## Independent Variable: 

- Huber (2012) measure of ethnicization.  
-- Uses surveys to determine cohesion of ethnic group voting behavior.  
-- Fearon (2003)'s list ethnic groups.  
-- Can be party-based or ethnic group based.   
-- Fractionalization or Polarization.  

- Loser Status:  
-- Determined by party affiliation in Afrobarometer.  
-- Use voting decision for CSES data.  

## Results H1: 

```{r, echo = FALSE, message=FALSE, warning=FALSE}

library(dotwhisker)
library(ggplot2)
library(dplyr)


my_data <- readRDS("total.rds")



library(lme4)
modela <- lmer(pol_trust ~ loser_party + nonpartisan + PVP + PVP*loser_party +
                education + employment + female + age + income + years_since_election + 
                ELF_ethnic + lngdp + polity2 + pr + lndm + country_isolation +
                (1 | country_wave), data=my_data)

model2a <- lmer(pol_trust ~ loser_party + nonpartisan + PVF + PVF*loser_party +
                education + employment + female + age + income + years_since_election + 
                ELF_ethnic + lngdp + polity2 + pr + lndm + country_isolation +
                (1 | country_wave), data=my_data)
Figure_1 <- dwplot(list(modela, model2a),
                   vline = geom_vline(xintercept = 0, colour = "grey60", linetype = 2)) %>% # plot line at zero _behind_ coefs
  relabel_predictors(c(loser_party = "Loser", 
                       nonpartisan = "Non-Partisan", 
                       PVP = "Party Ethnicization-P", 
                       PVF ="Party Ethnicization-F",
                       education = "Education",
                       employment = "Employed", 
                       female = "Female", 
                       age = "Age",
                       income = "Income",
                       years_since_election = "Years Since Election",
                       ELF_ethnic = "ELF", 
                       lngdp = "GDP log",
                       polity2 = "Polity",
                       pr = "PR",
                       lndm = "District Magnitude",
                       country_isolation = "Geographic Isolation" )) +
  theme_bw() + xlab("Coefficient Estimate") + ylab("") +
  geom_vline(xintercept = 0, colour = "grey60", linetype = 2) +
  ggtitle("Figure 1: Predicting Political Trust") +
  theme(plot.title = element_text(face="bold"),
        legend.position = c(0.007, 0.01),
        legend.justification = c(0, 0), 
        legend.background = element_rect(colour="grey80"),
        legend.title = element_blank() + 
          scale_color_grey(start = .3, end = .7,
                           name = "Models",
                           breaks = c(0,1),
                           labels = c("Polarization Model", "Fractionalization Model")))



Figure_1
```

## Results H2:  

```{r, echo = FALSE, message=FALSE, warning=FALSE}

my_data2 <- readRDS("cses_imd.rds")



model3 <- lmer(affpolwgt ~ Loser + PVP + nonpartisan +
                Education + employment + Female + Age + Income +
                ELF_ethnic + lngdp + polity2 + pr + lndm + country_isolation +
                (1 | Country_year), data=my_data2)

model4 <- lmer(affpolwgt ~ Loser + PVF + nonpartisan +
                 Education + employment + Female + Age + Income + 
                 ELF_ethnic + lngdp + polity2 + pr + lndm + country_isolation +
                 (1 | Country_year), data=my_data2)

Figure_2 <- dwplot(list(model3, model4),
                   vline = geom_vline(xintercept = 0, colour = "grey60", linetype = 2)) %>% # plot line at zero _behind_ coefs
  relabel_predictors(c(Loser= "Loser", 
                       nonpartisan = "Non-Partisan", 
                       PVP = "Party Ethnicization-P", 
                       PVF ="Party Ethnicization-F",
                       Education = "Education",
                       employment = "Employed", 
                       Female = "Female", 
                       Age = "Age",
                       Income = "Income",
                       ELF_ethnic = "ELF", 
                       lngdp = "GDP log",
                       polity2 = "Polity",
                       pr = "PR",
                       lndm = "District Magnitude",
                       country_isolation = "Geographic Isolation" )) +
  theme_bw() + xlab("Coefficient Estimate") + ylab("") +
  geom_vline(xintercept = 0, colour = "grey60", linetype = 2) +
  ggtitle("Figure 2: Predicting Affective Polarization") +
  theme(plot.title = element_text(face="bold"),
        legend.position = c(0.007, 0.01),
        legend.justification = c(0, 0), 
        legend.background = element_rect(colour="grey80"),
        legend.title = element_blank() + 
          scale_color_grey(start = .3, end = .7,
                           name = "Models",
                           breaks = c(0,1),
                           labels = c("Polarization Model", "Fractionalization Model")))

Figure_2
```

## Future Direction for Project:

- Examining the role of ethnic salience and sorting as conditioning factors.  
-- Requires expanding coverage of ethnicization measures.  

- Creating an ethnic sorting measure that incorporates (Individual-level ethnic salience, party dependence on ethnic groups as supporters (V-Party), and cohesion of the ethnic group).  

## Thank you

- Looking forward to any questions or suggestions



 

